---
title: "Salmon spatiotemporal fecundity estimates "
subtitle: "DIASPARA WP 2.2"
author: "Viktor Thunell" 
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  # docx:
  #   toc: true
  #   number-sections: true
  #   highlight-style: github
  html:
    code-fold: true
    code-summary: "Show code"
    page-layout: full
    embed-resources: true
    toc: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: true
reference-location: document
bibliography: diaspara.bib
---

## Introduction

Fish fecundity (the number of eggs per spawning female) is a key life history trait reflecting fitness resulting from evolutionary trade-offs and environmental conditions [@lambert2008]. In Atlantic salmon, *Salmo salar*, where semelparity is common (and generally the case in the North Atlantic while iteroparity is dominating in Baltic sea populations), fecundity becomes pivotal for individual reproductive potential and important for population dynamics [@crespi2002;@jonsson2017]. Fecundity is body size dependent with individual length being a suitable quantitative predictor  [@lambert2008]. Changes in the size of salmon returning to spawn can therefore be used to calculate a proxy for changes in their reproductive potential over time when temporally resolved fecundity data is missing, as in the case for the analysis of salmon fecundity in DIASPARA.

The working group on North Atlantic salmon (WGNAS) uses information on fecundity to estimate egg deposition for spawners that spent one or multiple winters in their marine phase (sea winters, SW). These estimate are stock unit specific and can vary over time. Here, knowledge of spatiotemporal changes in fecundity is key as the reproductive potential of many stock units assessed by WGNAS is at historically low levels [@wgnas2025]. 

The Assessment Working Group on Baltic Salmon and Trout (WGBAST) uses fecundity by sea winter age but this is currently not stock specific and is time invariant. Consequently, knowledge on changes in fecundity over time can help prioritize needs in future improvements of the stock assessment model [@wgbast2025]. 

To address these issues, we combine the spatial model for fecundity at length with the spatiotemporal model for length at age to produce spatiotemporal estimates of fecundity of salmon with varying number of winters spent at sea. A key assumption for this analyses is that individual fecundity is not subject to environmental variation nor evolutionary processes.

In detail, we propagate the uncertainty in spatial fecundity parameters $a_j$ and $b_j$ (for spatial unit $j$, see main text, section *Spatial variation in salmon fecundity - Model description*) to estimates of fecundity over time ($F_{y,j}$ in year $y$) by combining posterior draws of fecundity parameters $a$ and $b$ from the fecundity at length model with posterior draws of $l_{\mu}$ (predicted length at age) by spatial unit, cohort, age (growth year), and smolt age. Following the model formulation (main text), we calculate fecundity on the  $log$ scale based on mean centered $log$ length:

$$log(F_{y,j}) = a_j + b_j \times \log\left(\frac{l_{\mu}}{\bar{l_{\mu}}}\right)$$

As the model for length at age is smolt age specific, we predict spatiotemporal fecundity for smolt ages used in the assessment models. In the Baltic (for WGBAST) this is two year old smolts for Assessment Unit AU3 and three year old smolts in AU1 and AU2 (and we lack fecundity estimates for AU4-6). For the North Atlantic salmon and the WGNAS LCM model this is smolt ages 1-6, however fecundity is only predicted for existing smolt ages for a spatial unit.

```{r}
#| message: false
#| warning: false
#| cache: false
#| echo: false
#| output: false

# Load libraries, install if needed
pkgs <- c("tidyverse", "tidylog", "devtools","viridis","nls.multstart", "broom", "patchwork", "coda", "boot", "tidybayes","bayesplot", "nimble", "here", "purrr")

# remotes::install_github("nimble-dev/nimble", ref = "devel", subdir = "packages/nimble")

if(length(setdiff(pkgs,rownames(installed.packages()))) > 0){
    install.packages(setdiff(pkgs, rownames(installed.packages())), dependencies = T)
  }

invisible(lapply(pkgs, library, character.only = T))

options(ggplot2.continuous.colour = "viridis")
theme_set(theme_minimal())

# Set path
color_scheme_set("viridis")
home <- here::here()

## Read and define data
salfec <- readRDS(file = "~/gitProjects/diasp-lht/data/data-for-2-2/salmon-fec_2025-12-21.RData")

data.fec <- salfec %>%
  drop_na(spat.unit) %>%
  mutate(n.eggs = if_else(is.na(n.eggs.tot), n.eggs.str, n.eggs.tot),
         su = as.integer(factor(spat.unit)),
         length.lc = log(length_mm) - log(mean(length_mm)))

sallaa <- readRDS(file = "~/gitProjects/diasp-lht/data/data-for-2-2/salmon-laa_2025-12-21.RData") %>%
  rename(year = fi_year,
         site = sai_location,
         sea.age = sea_age_year,
         juv.age = juvenile_age_year,
         tot.age = tot_age_year,
         lat = fisa_y_4326,
         lon = fisa_x_4326) %>%
  #filter out tot.age = 0 individuals
  filter(!(sea.age == 0 & juv.age == 0)) %>%
  #filter out non-aged individuals (~90000 individuals)
  filter(!(is.na(sea.age) & is.na(juv.age))) %>%
  # removes smolt.age = 0:
  filter(!(age.type == "both" & juv.age == 0)) %>%
  mutate(hatch.year = year-tot.age) %>% # to filter data based o hatch year instead of year to predict growth pars before catch year
  filter(!is.na(juv.age), # removes age.type = "sea.only"
         !is.na(year),
         hatch.year > 1997)

data.biph <- sallaa %>%
  filter(!(is.na(sex) & age.type == "both")) %>% # removes 27 000 rows
  mutate(yday.may1 = if_else(is.na(date), 1, yday(date) - yday("2025-05-01")+1), 
         doy.dec = if_else(yday.may1 > 0, yday.may1, yday.may1+365)/365,
         tot.age.dec = if_else(yday.may1 > 0, tot.age+doy.dec, tot.age+doy.dec-1),
         smo.age = if_else(age.type == "both", juv.age, NA),
         sea.age = if_else(is.na(sea.age), 0, sea.age),
         g.year = trunc(tot.age.dec)+1, # + 1 to index age 0 individuals
         hatch.year.f = as.numeric(factor(hatch.year, levels = sort(unique(hatch.year)), labels = seq_along(sort(unique(hatch.year))))),
         year.f = as.integer(factor(year)),
         sex = as.integer(if_else(sex == "f", 1,2)),
         su = as.integer(factor(spat.unit))) %>%
  mutate(ind.id = row_number())

# "both" data (individuals caught at sea)
data.b <- data.biph %>%
  filter(!is.na(smo.age))

# "juve.age" data (immature individuals caught in freshwater)
data.j <- data.biph %>%
  filter(age.type == "juve.only") %>%
  rename(length_mm_j = length_mm)

```

## Combine length and fecundity MCMC samples
```{r}
#| message: false

# read fecundity at length samples 
fec.samples <- readRDS(paste0(home,"/models_salmon/samples/fec_samples_2026-01-15.RData"))$samples %>%
  spread_draws(a[su],b[su]) %>%
  # remove draws aove 1500 to match number with laa samples
  filter(.draw <= 1500) %>%
  left_join(data.fec %>% distinct(su, spat.unit)) %>%
  ungroup() %>%
  select(-su,-.iteration,-.chain )

# read length at age mcmc samples 
laa.samples <- readRDS(paste0(home,"/models_salmon/samples/salaa_samples_2025-12-22.RData"))$samples

masu <- data.biph %>%
  filter(!is.na(smo.age)) %>%
  summarise(ma = max(g.year), .by = su) %>%
  arrange(su)

# select the expected value parameter l.mu
node.sub <- grep("l.mu", colnames(laa.samples[[1]]), value = TRUE)
laa.samples_2 <- laa.samples[, node.sub[], drop = FALSE] %>%
  spread_draws(l.mu[su,smo.age,coh,sex,gy], sep = ",") %>%
  # remove males and draw 1501 and 1 (a few NA l.mu:s there)
  filter(sex == 1,
         gy %in% c(smo.age+2, smo.age+3, smo.age+4, smo.age+5, smo.age+6, smo.age+7), 
         .draw <= 1500 & .draw > 1
         ) %>%
  left_join(data.biph %>% distinct(su, spat.unit)) %>%
  # filter out the spatial units for which we have fecundity estimates
  filter(spat.unit %in% unique(fec.samples$spat.unit)) %>%
  # filter out the most common smolt age per spatial unit and maximum ages
  filter(#smo.age == mosa[su,2],
         gy <= masu[su,2])  %>%
  ungroup() %>%
  select(-sex,-su,-.iteration,-.chain )

```

## Calculate fecundity per MCMC draw
```{r}
#| message: false

lf.samples <- laa.samples_2 %>%
  # join in mcmc draws of a and b from fecundity model
  left_join(fec.samples,  by = c("spat.unit",".draw")) %>%
  mutate(length.lc = log(l.mu) - log(mean(l.mu)),
         fec = exp(a + b * length.lc),
         year = coh + gy + 1998) %>%
  rename(length = l.mu)
```

## Results

Changes in fecundity over time is presented for 1 and 2 SW spawners with posterior uncertainty in @fig-salft1 and @fig-salft2, for 3-5SW Baltic spawners in @fig-salft3, and for all ages spawners that has occurred within the spatial units in the analysed time frame (@fig-salft4,without posterior uncertainty). The dependence on smolt age in our biphasic growth model for North Atlantic salmon, resulting in variation in fecundity is shown in @fig-salft3b.

The analyses suggest considerable regional differences in the change in fecundity over time where in the Baltic sea, fecundity would have increased in the analysed time frame, while it is the opposite for the North Atlantic (France, Ireland and Sweden). Here, France and Sweden share a similar temporal trend where recent years 

Note that the temporal pattern in this analysis reflect changes in length for fish returning to spawn and spatial-unit-specific length-fecundity relationships (see main text). The length at age of salmon here follows a von Bertalanffy type growth trajectory and differences between the ages of salmon within a spatial unit and year will differ with an offset given by their growth trajectory. This effect is dependent on the temporal trend in the von Bertalanffy growth coefficient $k$ and as growth levels off with age, changes in $k$ with time has stronger effects on yong spawners compared to older spawners (compare e.g. 1 and 3 SW fecundity in Ireland_west, @fig-salft4).

Given the considerable change in fecundity resulting from changes in length in the North Atlantic spatial units, it should have large consequences for the reproductive capacity of the stocks given that common age at spawning remains constant. 

::: {#fig-salft1}
```{r}
#| message: false

lf.samples %>%
  filter(
    # WGBAST assume 3 year old smolts in AU1 or 2, and then 2 year old smolts in AU 3 and 4.  
    !(spat.unit %in% c("Baltic.sea:AU-1","Baltic.sea:AU-2") & !smo.age == 3),
    !(spat.unit %in% c("Baltic.sea:AU-3","Baltic.sea:AU-4") & !smo.age == 2),
    gy == smo.age+2, # 1SW
         year < 2025,
         #!spat.unit %in% c("Baltic.sea:AU-2","Baltic.sea:AU-3")
         ) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(.75))) +
  labs(title = "1SW")
```

Fecundity (number of eggs) of 1SW salmon over time by spatial unit (both North Atlantic and Baltic). In the Baltic sea, smolt age 2 in shown for Assessment Unit AU3 and smolt age 3 in AU1 and AU2, for the North Atlantic salmon, fecundity is marginalised over smolt ages. Boxplots summarise the posterior draws for each year and spatial unit  where the center line is the median, boxes and whiskers show the IQR and 1.5 times IQR and outliers are excluded.
:::

::: {#fig-salft2}
```{r}
lf.samples %>%
  filter(# WGBAST assume 3 year old smolts in AU1 or 2, and then 2 year old smolts in AU 3 and 4.  
    !(spat.unit %in% c("Baltic.sea:AU-1","Baltic.sea:AU-2") & !smo.age == 3),
    !(spat.unit %in% c("Baltic.sea:AU-3","Baltic.sea:AU-4") & !smo.age == 2),
    gy == smo.age+3,
    # remove years (2025) without data
    year < 2025) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 90, hjust = 1, size = rel(.75))) +
  labs(title = "2SW")
```

Fecundity (number of eggs) of 2SW salmon over time by spatial unit (both North Atlantic and Baltic). In the Baltic sea, smolt age 2 in shown for Assessment Unit AU3 and smolt age 3 in AU1 and AU2, for the North Atlantic salmon, fecundity is marginalised over smolt ages. Boxplots summarise the posterior draws for each year and spatial unit  where the center line is the median, boxes and whiskers show the IQR and 1.5 times IQR and outliers are excluded.
:::

::: {#fig-salft3}
```{r}
lf.samples %>%
  filter(case_when(
    # WGBAST assume 3 year old smolts in AU1 or 2, and then 2 year old smolts in AU 3 and 4.  
    spat.unit %in% c("Baltic.sea:AU-1","Baltic.sea:AU-2") ~ smo.age == 3,
    spat.unit %in% c("Baltic.sea:AU-3","Baltic.sea:AU-4") ~ smo.age == 2),
    # filter out 3, 4 and 5 SW
    gy == (smo.age + 4) | gy == (smo.age + 5) | gy == (smo.age + 6),
    #gy %in% c(smo.age+4,smo.age+5,smo.age+6)) %>%
   year < 2025) %>%
  mutate(SW = paste0(gy-smo.age-1,"SW")) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_grid(spat.unit~SW, scales = "free") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 90, hjust = 1, size = rel(.75))) 
```

Fecundity (number of eggs) of 3-5SW salmon over time by spatial unit in the Baltic sea for smolt age 2 in Assessment Unit AU3 and smolt age 3 in AU1 and AU2. Boxplots summarise the posterior draws for each year and spatial unit  where the center line is the median, boxes and whiskers show the IQR and 1.5 times IQR and outliers are excluded.
:::

::: {#fig-salft3b}
```{r}
lf.samples %>%
  filter(# WGBAST assume 3 year old smolts in AU1 or 2, and then 2 year old smolts in AU 3 and 4.  
    !str_detect(spat.unit,"AU"),
    gy == (smo.age + 2) | gy == (smo.age + 3),
    # remove years (2025) without data
    year < 2025) %>%
  mutate(SW = paste0(gy-smo.age-1,"SW")) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year), color = SW),
               outliers = FALSE) +
  facet_grid(smo.age~spat.unit, scales = "free_y") +
  scale_x_discrete(breaks = seq(2002, 2024, by = 2)) +
  theme_minimal() +
  theme(axis.text = element_text(angle = 90, hjust = 1, size = rel(.75))) 
```

Fecundity (number of eggs) of 1 and 2SW salmon over time by North Atlantic spatial units (columns) and smolt age 1-5 (rows). Boxplots summarise the posterior draws for each year and spatial unit  where the center line is the median, boxes and whiskers show the IQR and 1.5 times IQR and outliers are excluded.
:::

::: {#fig-salft4}
```{r}
lf.samples %>%
  group_by(year,gy,spat.unit) %>%
  median_qi(fec) %>%
  # remove years (2025) without data
  filter(year < 2025,
         !fec == 0) %>%
  ggplot() +
  # minus 2 because gy = 2 is 1SW (smo.age is age = 0 and gy is age+1)
  geom_line(aes(y = fec, x = year, color = factor(gy-2))) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = rel(.75))) +
  labs(color = "#sea winters", y = "#eggs")
```

Fecundity (number of eggs) of salmon over time by spatial unit and number of years spent at sea (sea winters, color scale). Predictions are shown for all sea winters salmon in a spatial unit existing over the time period despite some being rare. In the Baltic sea, smolt age 2 in shown for Assessment Unit AU3 and smolt age 3 in AU1 and AU2, for the North Atlantic salmon, fecundity is marginalised over smolt ages.
:::
