---
title: "Salmon spatiotemporal fecundity estimates "
subtitle: "DIASPARA WP 2.2"
author: "Viktor Thunell" 
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    page-layout: full
    embed-resources: true
    toc: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: true
---

## Introduction 
- Combine spatial FAL with spatiotemporal LAA to construct spatiotemporal FAL
- Propagate the uncertainty in fecundity parameters $a$ and $b$ to estimates of fecundity over time
  * Create samples of individual fecundity estimates
    * from sampling the distributions of the allometric parameters $a$ and $b$
    * and using estimates of $l_{\mu}$ in $a_{s_i}l_{\mu}^{b_{s_i}}$
    * (used deterministic $l_{\mu}$ estimates for convenience)

## Load libraries

```{r}
#| message: false
#| warning: false
#| cache: false

# Load libraries, install if needed
pkgs <- c("tidyverse", "tidylog", "devtools","viridis","nls.multstart", "broom", "patchwork", "coda", "boot", "tidybayes","bayesplot", "nimble", "here", "purrr")

# remotes::install_github("nimble-dev/nimble", ref = "devel", subdir = "packages/nimble")

if(length(setdiff(pkgs,rownames(installed.packages()))) > 0){
    install.packages(setdiff(pkgs, rownames(installed.packages())), dependencies = T)
  }

invisible(lapply(pkgs, library, character.only = T))

options(ggplot2.continuous.colour = "viridis")
theme_set(theme_minimal())

# Set path
color_scheme_set("viridis")
home <- here::here()
```


```{r}
## l.mu and Fecundity over time

### l.mu

```{r}
# Function for calculating l.mu
calc_l_mu <- nimbleFunction(
  run = function(
    lb_mu = double(1), 
    g = double(2), 
    log_p_year = double(2),
    su = double(1),
    age = double(1),
    smo_age = double(1),
    hatch_year_f = double(1),
    growth_year = double(1),
    doy_dec = double(1),
    max_age = double(0)
  ) {
    returnType(double(2))
    
    nind <- length(su)
    nyears <- dim(g)[1]
    nsu <- dim(g)[2]
    
    # initialize output with sentinel (-999)
    l_mu <- matrix(-999, nrow = max_age + 1, ncol = nind)
    
    for (i in 1:nind) {
      su_i <- su[i]
      hyf_i <- hatch_year_f[i]
      ag_i <- age[i]
      smo_i <- smo_age[i]
      gy_i <- growth_year[i]
      doy_i <- doy_dec[i]
      
      # initial 0+ prediction
      l_prev <- lb_mu[su_i] +
        (1.0 - step(1.0 - (ag_i + 0.001))) *
        (exp(g[hyf_i, su_i]) + step(1.0 - (ag_i + 0.001)) * exp(g[hyf_i, su_i]) * doy_i)
      
      l_mu[1, i] <- l_prev
      
      if (gy_i >= 1) {
        for (a in 2:(gy_i + 1)) {
          year_idx <- hyf_i + (a - 1)
          outside <- step(1.0 - year_idx) + step(year_idx - nyears)
          
          if (outside == 0.0) {
            pre_smolt <- step(smo_i - (a - 0.999))
            post_smolt <- 1.0 - pre_smolt
            before_age <- 1.0 - step((a - 1) - (ag_i + 0.001))
            
            g_exp <- exp(g[year_idx, su_i])
            linf_exp <- exp(log_p_year[year_idx, su_i])
            k_exp <- exp(log_p_year[year_idx, nsu + su_i])
            
            delta_len <- before_age * (
              pre_smolt * g_exp +
                post_smolt * ((linf_exp - l_prev) * (1.0 - exp(-k_exp)))
            ) +
              (1.0 - before_age) * (
                pre_smolt * g_exp * doy_i +
                  post_smolt * ((linf_exp - l_prev) * (1.0 - exp(-k_exp * doy_i)))
              )
            
            l_new <- l_prev + delta_len
            l_prev <- l_new
            l_mu[a, i] <- l_new
          } else {
            l_mu[a, i] <- l_prev
          }
        } # end loop a
      } # end if gy_i >= 1
    } # end loop i
    
    return(l_mu)
  }
)

recompute_lmu <- function(samples, consts, calc_l_mu_c, n_draws) {
  message("Preparing to recompute l.mu from posterior samples...")

  nyear <- consts$nyear
  nsu   <- consts$nsu
  nobs  <- length(consts$su)
  max_age <- max(consts$growth.year, na.rm = TRUE)

  # extract parameter arrays
  lb_mu_samps <- extract_node_samples(samples, "^lb.mu\\[", c(nsu))
  g_samps     <- extract_node_samples(samples, "^g\\[", c(nyear, nsu))
  logp_samps  <- extract_node_samples(samples, "^log_p.year\\[", c(nyear, consts$npars))
  smo_age_samps <- extract_node_samples(samples, "^smo.age\\[", c(consts$nobs))

  n_draws <- min(n_draws, dim(lb_mu_samps)[1])
  message(sprintf("Running %d posterior draws...", n_draws))
  
  lmu_list <- vector("list", n_draws)
  
  for (i in seq_len(n_draws)) {
    if (i %% 50 == 0) message("  • Draw ", i, "/", n_draws)

    lb_mu_i   <- as.numeric(lb_mu_samps[i, , drop = TRUE])
    g_i       <- g_samps[i, , , drop = TRUE]
    logp_i    <- logp_samps[i, , , drop = TRUE]
    smo_age_i <- as.numeric(smo_age_samps[i, , drop = TRUE])
    
    lmu_i <- calc_l_mu_c(
      lb_mu        = lb_mu_i,
      g            = g_i,
      log_p_year   = logp_i,
      smo_age      = smo_age_i,
      su           = consts$su,
      age          = consts$age,
      hatch_year_f = consts$hatch.year.f,
      growth_year  = consts$growth.year,
      doy_dec      = consts$doy.dec,
      max_age      = max_age
    )

    # replace sentinel with NA
    lmu_i[lmu_i == -999] <- NA_real_
    lmu_list[[i]] <- lmu_i
  }

  message("Finished recomputing all deterministic l.mu values.")
  return(lmu_list)
}

extract_node_samples <- function(samples, pattern, dims) {
  cols <- grep(pattern, colnames(samples), value = TRUE)
  if (length(cols) == 0)
    stop("No columns found for pattern: ", pattern)

  arr <- as.matrix(samples[, cols, drop = FALSE])
  ndraws <- nrow(arr)
  dims <- as.integer(unlist(dims))
  ncols <- ncol(arr)
  expected <- prod(dims)

  # Handle mismatched column counts gracefully
  if (expected != ncols) {
    warning(sprintf(
      "Dimension mismatch for pattern '%s': expected %d, got %d. Adjusting dims automatically.",
      pattern, expected, ncols
    ))

    if (length(dims) == 2) {
      inferred_nyear <- ncols / dims[2]
      dims <- c(as.integer(inferred_nyear), dims[2])
    } else if (length(dims) == 3) {
      inferred_nyear <- ncols / (dims[2] * dims[3])
      dims <- c(as.integer(inferred_nyear), dims[2], dims[3])
    } else if (length(dims) == 1) {
      dims <- ncols
    }
    message(" → Inferred dims for ", pattern, ": ", paste(dims, collapse = " x "))
  }

  # Always produce an array with draw dimension first
  full_dims <- c(ndraws, dims)
  out <- array(as.numeric(arr), dim = full_dims)

  # Guarantee minimum dimension of 2 (for single-node cases)
  if (length(dim(out)) == 1) {
    dim(out) <- c(ndraws, length(cols))
  }

  return(out)
}

calc_l_mu_c <- compileNimble(calc_l_mu)
lmu_list <- recompute_lmu(as.data.frame(biph.samples), consts, calc_l_mu_c, n_draws = nrow(biph.samples))
str(lmu_list) # [years, ind.id]

```

## Option 2

```{r}

lb_mu_med <- extract_node_samples(biph.samples, "^lb.mu\\[", c(consts$nsu)) |>
  apply(2, median, na.rm = TRUE)

g_med <- extract_node_samples(biph.samples, "^g\\[", c(consts$nyear, consts$nsu)) |>
  apply(c(2, 3), median, na.rm = TRUE)

logp_med <- extract_node_samples(biph.samples, "^log_p.year\\[", c(consts$nyear, consts$npars)) |>
  apply(c(2, 3), median, na.rm = TRUE)


# Ensure correct integer smolt ages (if continuous, round)
# mean_smo_df <- mean_smo_df %>%
#   mutate(m_smo_age = round(mean_smo_age))
m_smo_df <- data.biph %>%
  summarise(msmo = median(smo.age, na.rm = TRUE), .by = su)

su_vec   <- rep(1:consts$nsu, each = consts$nyear)
year_vec <- rep(1:consts$nyear, times = consts$nsu)
nind_sim <- length(su_vec)

# Join smolt age by spatial unit
smo_age_sim <- m_smo_df$msmo[match(su_vec, m_smo_df$su)]

# Other deterministic covariates
age_sim <- 6 # arbitrary upper age (6 years) for all
growth_year_sim <- age_sim
doy_dec_sim <- 1.0  # end-of-year condition (full year of growth)
hatch_year_f_sim <- pmax(1, year_vec - growth_year_sim + 1)  # ensure valid index
max_age <- growth_year_sim

lmu_sim <- calc_l_mu_c(
  lb_mu        = lb_mu_med,
  g            = g_med,
  log_p_year   = logp_med,
  su           = su_vec,
  age          = rep(age_sim, nind_sim),
  smo_age      = smo_age_sim,
  hatch_year_f = hatch_year_f_sim,
  growth_year  = rep(growth_year_sim, nind_sim),
  doy_dec      = rep(doy_dec_sim, nind_sim),
  max_age      = max_age
)

lmu_sim[lmu_sim == -999] <- NA_real_


lmu_df <- as.data.frame(lmu_sim) |>
  mutate(age_idx = row_number()) |>
  pivot_longer(-age_idx, names_to = "ind", values_to = "length_mm") |>
  mutate(ind = as.integer(str_remove(ind, "V")),
         su = su_vec[ind],
         year = year_vec[ind],
         smo_age = smo_age_sim[ind])

lmu_df %>%
  left_join(data.biph %>% distinct(spat.unit, su)) %>%
  filter(year %in% sample(1:26,10)) %>%
  ggplot(aes(x = age_idx - 1, y = length_mm, group = interaction(spat.unit, year), color = factor(year)), show.legend = FALSE) +
  geom_line(alpha = 0.7) +
  facet_wrap(~spat.unit) +
  geom_vline(aes(xintercept = smo_age), linetype = 2, color = "grey40", alpha = 0.5) +
  labs(x = "age", y = "predicted length") +
  theme_minimal() 

# ggplot(aes(age_idx - 1, length_mm, group = ind, color = spat.unit)) +
#   
#   geom_line() +
#   #geom_vline(aes(xintercept = smo_age), linetype = 2, color = "grey50", alpha = 0.4) +
#   labs(x = "age",y = "predicted length [mm]") +
#   facet_wrap(~spat.unit) +
#   theme_minimal() +
#   theme(panel.grid = element_blank())
```

### f.mu
```{r}
compute_fmu_summary_from_lmu_df <- function(lmu_df, fec_dpars, ndraws = 2000) {
  # Ensure required columns exist
  stopifnot(all(c("spat.unit", "year", "age_idx", "length_mm") %in% names(lmu_df)))
  stopifnot(all(c("spat.unit", "shfa", "rafa", "mufb", "sdfb") %in% names(fec_dpars)))
  
  nsu <- nrow(fec_dpars)
  
  message("Simulating fecundity using uncertainty in a and b parameters...")
  
  # Preallocate output list
  out_list <- vector("list", nsu)
  
  for (s in seq_len(nsu)) {
    su_name <- fec_dpars$spat.unit[s]
    
    # Draw random samples for a and b
    fa_draws <- rgamma(ndraws, shape = fec_dpars$shfa[s], rate = fec_dpars$rafa[s])
    fb_draws <- rnorm(ndraws, mean = fec_dpars$mufb[s], sd = fec_dpars$sdfb[s])
    
    # Subset the deterministic lengths for this spatial unit
    lsu <- lmu_df %>% filter(spat.unit == su_name)
    
    # Calculate predicted fecundity for each sample
    fmu_samples <- replicate(
      ndraws,
      lsu$length_mm^sample(fb_draws, 1) * sample(fa_draws, 1)
    )
    
    # Summarize posterior predictive distribution per row (age × year)
    lsu_summary <- lsu %>%
      mutate(
        fmu_med = apply(fmu_samples, 1, median, na.rm = TRUE),
        fmu_low = apply(fmu_samples, 1, quantile, probs = 0.025, na.rm = TRUE),
        fmu_high = apply(fmu_samples, 1, quantile, probs = 0.975, na.rm = TRUE)
      )
    
    out_list[[s]] <- lsu_summary
  }
  
  result <- bind_rows(out_list)
  
  message("✅ Finished computing f.mu summaries across spatial units.")
  return(result)
}

#fec_samples_df <- do.call(rbind, fec.samples$samples) |> as.data.frame()

#lmu_df <- lmu_df %>% left_join(data.biph %>% distinct(su,spat.unit)) %>% select(-su)
fmu_summary <- compute_fmu_summary_from_lmu_df(
  lmu_df = lmu_df %>% left_join(data.biph %>% distinct(su,spat.unit)),
  fec_dpars = fec.dpars,
  ndraws = 2000
)
lmu_fec <- lmu_df %>%
  left_join(fmu_summary %>% select(spat.unit, su, year, age_idx, fmu_med, fmu_low, fmu_high),
            by = c("spat.unit", "year", "age_idx"))

saveRDS(lmu_fec, file = paste0(home,"/data/lmu_fec_1104.RData"))
```

### Reshape l.mu and f.mu:s
```{r}
# but first save lmu and fmu.
#flmu_list <- list(fmu_list,lmu_list)
#saveRDS(flmu_list, file = paste0(home,"/data/flmu_list_1102.RData"))


flmu_chunk <- function(flmu_list_chunk) {
  purrr::imap_dfr(flmu_list_chunk, function(mat, draw_id) {
    tibble::tibble(
      draw = as.integer(draw_id),
      age  = rep(seq_len(nrow(mat)), times = ncol(mat)),
      ind  = rep(seq_len(ncol(mat)), each = nrow(mat)),
      value = as.vector(mat)
    )
  }) |>
    dplyr::filter(!is.na(value)) |>
    dplyr::group_by(age, ind) |>
    dplyr::summarise(
      q50 = median(value),
      q025 = quantile(value, 0.025),
      q975 = quantile(value, 0.975),
      .groups = "drop"
    )
}

flmu_mqi <- function(flmu_list, chunk_size = 250) {
  n_draws <- length(flmu_list)
  chunks <- split(seq_len(n_draws), ceiling(seq_len(n_draws) / chunk_size))
  
  chunk_summaries <- purrr::map(chunks, function(idx) {
    summarise_lmu_chunk(flmu_list[idx])
  })
  
  # Combine across chunks by taking median of medians, etc.
  dplyr::bind_rows(chunk_summaries) |>
    dplyr::group_by(age, ind) |>
    dplyr::summarise(
      median = median(q50, na.rm = TRUE),
      lower  = median(q025, na.rm = TRUE),
      upper  = median(q975, na.rm = TRUE),
      .groups = "drop"
    )
}

lmu_df <- flmu_mqi(flmu_list)

#f.mu samps
fmu_df <- flmu_mqi(fmu_list)

flmu_df <- bind_rows(lmu_df %>% mutate(variable = "l.mu"), lmu_df %>% mutate(variable = "f.mu"))
saveRDS(flmu_df, file = paste0(home,"/data/flmu_df_1102.RData"))

```

approach 1, unfinished but using growth parameters estimates to predict l.mu
```{r}
growth.samples <- biph.samples[1:2000,]

growth_pars1 <- growth.samples %>%
  spread_draws(log_p.year[yr,sup], sep = ",") %>%
  filter(sup %in% 1:12) %>% # first 12 sup:s are Linf 
  rename(Linf = log_p.year,
         su = sup) %>%
  ungroup() %>%
  #filter(.chain == 1) %>%
  select(-.iteration,-.chain)

growth_pars2 <- growth.samples %>%
  spread_draws(log_p.year[yr,sup], sep = ",") %>%
  filter(sup %in% 13:24) %>% 
  mutate(su = sup - 12) %>% 
  rename(k = log_p.year) %>%
  ungroup() %>%
  #filter(.chain == 1) %>%
  select(-.iteration,-.chain,-sup)

growth_pars3 <- growth.samples %>%
  spread_draws(lb.mu[su], g[yr,su], sep = ",") %>%
  ungroup() %>%
  #filter(.chain == 1) %>%
  select(-.iteration,-.chain)

#su:s in growth data
sun <- data.biph %>% distinct(su, spat.unit)
growth.pars <- left_join(growth_pars1, growth_pars2) %>%
  left_join(growth_pars3) %>%
  left_join(sun) %>%
  select(-su)

# smolt age samples
sages <- growth.samples %>%
  spread_draws(smo.age[i]) %>%
  ungroup() %>%
  #filter(.chain == 1) %>%
  select(-.iteration,-.chain)

#fec
samplesf <- fec.samp$samples$chain1[1:2000,]
sunf <- data.fec %>%
  distinct(spat.unit,su)

fec_pars <- samplesf %>%
  spread_draws(a[su], b[su],sep = ",") %>%
  select(-.iteration,-.chain) %>%
  ungroup() %>%
  left_join(sunf) %>%
  select(-su)

# Step 3: join the two draw frames (assuming independence)
# This matches draw i from each posterior; you can also do cartesian join if needed
# joint_draws <- bind_cols(
#   growth_pars %>% arrange(.draw),
#   fec_pars %>% arrange(.draw) %>% select(-.draw)
# )

joint_draws <- 
  growth.pars %>% 
  filter(spat.unit %in% sunf$spat.unit) %>%
  left_join(fec_pars %>% filter(spat.unit %in% sun$spat.unit),
            by = join_by(.draw,spat.unit))
  
joint_draws_age <- joint_draws %>%
  expand_grid(age = 1:10)
  
# Step 4: Generate predicted length and fecundity per draw
# Replace with your actual function that rebuilds length for given parameters
length_pred <- function(lb_mu, par, g, age, doy_dec = 1) {
  # Example placeholder: VB increment
  lb_mu + (exp(Linf) - lb_mu) * (1 - exp(-exp(k) * age * doy_dec))
}

fec_pred <- function(L, a, b) a * L^b

# Suppose you want predictions for su = 1..nsu, years = vector of interest
preds <- 
  expand_grid(su = 1:nsu,
              year = years_vec,
              draw = 1:nrow(joint_draws)) %>%
  # extract parameters per draw/spatial unit
  mutate(lb = map2_dbl(draw, su, ~ joint_draws[[paste0("lb.mu[", .y, "]")]][.x]),
         a  = map2_dbl(draw, su, ~ joint_draws[[paste0("a[", .y, "]")]][.x]),
         b  = map2_dbl(draw, su, ~ joint_draws[[paste0("b[", .y, "]")]][.x]),
         # for brevity, pick one age for fecundity (say age=2)
         L  = predict_length(lb, 
                             par = c(joint_draws[[paste0("log_p.year[", year, ",", su, "]")]][draw]),
                             g   = joint_draws[[paste0("g[", year, ",", su, "]")]][draw],
                             age = 2),
         fec = predict_fecundity(L, a, b))

# Step 5: summarize with tidybayes
fec_summary <- preds %>%
  group_by(su, year) %>%
  median_qi(fec, .width = c(.5, .8, .95))

fec_summary
```

```



## Fecundity over time 
:::: columns
::: {.column width="50%"}
  
:::
::: {.column width="450%"}
```{r fig.width=5, fig.height=5, out.width="90%"}
lmu_fec %>%
  rename(age = age_idx) %>%
  filter(age > smo_age) %>%
  filter(year %in% sample(1:26,10)) %>%
  ggplot(aes(x = age - 1, y = length_mm, group = interaction(spat.unit, year), color = factor(year)), show.legend = FALSE) +
  geom_line(alpha = 0.7) +
  facet_wrap(~spat.unit) +
  geom_vline(aes(xintercept = smo_age), linetype = 2, color = "grey40", alpha = 0.5) +
  labs(x = "age", y = "predicted length") +
  theme_minimal() 
```
:::
::::
## Fecundity over time 
```{r fig.width=7, fig.height=4, out.width="150%"}
lmu_fec %>%
  rename(age = age_idx) %>%
  filter(age > smo_age) %>%
  filter(age == smo_age+2) %>%# 1 SW
  #filter(variable == "f.mu") %>%
  ggplot(aes(year, fmu_med)) +
  geom_line() +
  geom_errorbar(aes(ymin = fmu_low, ymax = fmu_high), alpha = 0.3) + 
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  labs(title = "2SW", subtitle = "median & 95% cred. intrval", y = "#eggs") 

```
## Fecundity over time 
```{r fig.width=7, fig.height=4, out.width="150%"}
lmu_fec %>%
  rename(age = age_idx) %>%
  filter(age > smo_age) %>%
  filter(age == smo_age+2) %>%# 1 SW
  #filter(variable == "f.mu") %>%
  ggplot(aes(year, fmu_med)) +
  geom_line() +
  #geom_errorbar(aes(ymin = fmu_low, ymax = fmu_high), alpha = 0.3) + 
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  labs(title = "2SW", subtitle = "median", y = "#eggs") 

```