---
title: "Salmon spatiotemporal fecundity estimates "
subtitle: "DIASPARA WP 2.2"
author: "Viktor Thunell" 
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    page-layout: full
    embed-resources: true
    toc: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: true
---

## Introduction 

- Combine the model for spatial fecundity model with spatiotemporal model for length at age to produce spatiotemporal estimates of fecundity of 1,2 and 3 SW salmon.
- Propagate the uncertainty in fecundity parameters $a_s$ and $b_$ to estimates of fecundity over time by
  * Combining posterior draws of fecundity parameters $a$ and $b$ (by spatial unit) from fecundity at length model
  * with posterior draws of $l_{\mu}$ by spatial unit, cohort, age (growth year), and smolt age
  * and calculate $log(F_{y,j}) = a_j + b_j \times \log\left(\frac{L_i}{\bar{L}}\right)$

## Load libraries

```{r}
#| message: false
#| warning: false
#| cache: false

# Load libraries, install if needed
pkgs <- c("tidyverse", "tidylog", "devtools","viridis","nls.multstart", "broom", "patchwork", "coda", "boot", "tidybayes","bayesplot", "nimble", "here", "purrr")

# remotes::install_github("nimble-dev/nimble", ref = "devel", subdir = "packages/nimble")

if(length(setdiff(pkgs,rownames(installed.packages()))) > 0){
    install.packages(setdiff(pkgs, rownames(installed.packages())), dependencies = T)
  }

invisible(lapply(pkgs, library, character.only = T))

options(ggplot2.continuous.colour = "viridis")
theme_set(theme_minimal())

# Set path
color_scheme_set("viridis")
home <- here::here()
```

## Read and define data

```{r}
#| message: false
#| warning: false
#| cache: false

salfec <- readRDS(file = "~/gitProjects/diasp-lht/data/data-for-2-2/salmon-fec_2025-12-21.RData")

data.fec <- salfec %>%
  drop_na(spat.unit) %>%
  mutate(n.eggs = if_else(is.na(n.eggs.tot), n.eggs.str, n.eggs.tot),
         su = as.integer(factor(spat.unit)),
         length.lc = log(length_mm) - log(mean(length_mm)))

sallaa <- readRDS(file = "~/gitProjects/diasp-lht/data/data-for-2-2/salmon-laa_2025-12-21.RData") %>%
  rename(year = fi_year,
         site = sai_location,
         sea.age = sea_age_year,
         juv.age = juvenile_age_year,
         tot.age = tot_age_year,
         lat = fisa_y_4326,
         lon = fisa_x_4326) %>%
  #filter out tot.age = 0 individuals
  filter(!(sea.age == 0 & juv.age == 0)) %>%
  #filter out non-aged individuals (~90000 individuals)
  filter(!(is.na(sea.age) & is.na(juv.age))) %>%
  # removes smolt.age = 0:
  filter(!(age.type == "both" & juv.age == 0)) %>%
  mutate(hatch.year = year-tot.age) %>% # to filter data based o hatch year instead of year to predict growth pars before catch year
  filter(!is.na(juv.age), # removes age.type = "sea.only"
         !is.na(year),
         hatch.year > 1997)

data.biph <- sallaa %>%
  filter(!(is.na(sex) & age.type == "both")) %>% # removes 27 000 rows
  mutate(yday.may1 = if_else(is.na(date), 1, yday(date) - yday("2025-05-01")+1), 
         doy.dec = if_else(yday.may1 > 0, yday.may1, yday.may1+365)/365,
         tot.age.dec = if_else(yday.may1 > 0, tot.age+doy.dec, tot.age+doy.dec-1),
         smo.age = if_else(age.type == "both", juv.age, NA),
         sea.age = if_else(is.na(sea.age), 0, sea.age),
         g.year = trunc(tot.age.dec)+1, # + 1 to index age 0 individuals
         hatch.year.f = as.numeric(factor(hatch.year, levels = sort(unique(hatch.year)), labels = seq_along(sort(unique(hatch.year))))),
         year.f = as.integer(factor(year)),
         sex = as.integer(if_else(sex == "f", 1,2)),
         su = as.integer(factor(spat.unit))) %>%
  mutate(ind.id = row_number())

# "both" data (individuals caught at sea)
data.b <- data.biph %>%
  filter(!is.na(smo.age))

# "juve.age" data (immature individuals caught in freshwater)
data.j <- data.biph %>%
  filter(age.type == "juve.only") %>%
  rename(length_mm_j = length_mm)

```

## Combine length and fecundity samples
```{r}
laa.samples <- readRDS(paste0(home,"/models_salmon/samples/salaa_samples_2025-12-22.RData"))$samples

fec.samples <- readRDS(paste0(home,"/models_salmon/samples/fec_samples_2026-01-15.RData"))$samples %>%
  spread_draws(a[su],b[su]) %>%
  filter(.draw <= 1500) %>%
  left_join(data.fec %>% distinct(su, spat.unit)) %>%
  ungroup() %>%
  select(-su,-.iteration,-.chain )

node.sub <- grep("l.mu", colnames(laa.samples[[1]]), value = TRUE)
laa.samples_2 <- laa.samples[, node.sub[], drop = FALSE] %>%
  spread_draws(l.mu[su,smo.age,coh,sex,gy], sep = ",") %>%
  filter(sex == 1,
         gy %in% c(smo.age+2, smo.age+3, smo.age+4)  # for 1-3 SW (gy is age+1)
         ) %>%
  left_join(data.biph %>% distinct(su, spat.unit)) %>%
  filter(spat.unit %in% unique(fec.samples$spat.unit)) %>%
  ungroup() %>%
  select(-sex,-su,-.iteration,-.chain ) %>%
  filter(!is.na(l.mu)) # .draw 1 and 1501 is NA so remove these

lf.samples <- laa.samples_2 %>%
  left_join(fec.samples,  by = c("spat.unit",".draw")) %>%
  filter(!is.na(l.mu)) %>%
  mutate(length.lc = log(l.mu) - log(mean(l.mu)),
         fec = exp(a + b * length.lc),
         year = coh + gy + 1998) %>%
  rename(length = l.mu)

```

```{r}

lf.samples %>%
  filter(gy == smo.age+2,
         year < 2025,
         !spat.unit %in% c("Baltic.sea:AU-2","Baltic.sea:AU-3")) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = rel(.75))) +
  labs(title = "1SW")

lf.samples %>%
  filter(gy == smo.age+3,
         year < 2025,
         !spat.unit %in% c("Baltic.sea:AU-2","Baltic.sea:AU-3")) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = rel(.75))) +
  labs(title = "2SW")

lf.samples %>%
  filter(gy == smo.age+4,
         year < 2025,
         !spat.unit %in% c("Baltic.sea:AU-2","Baltic.sea:AU-3")) %>%
  ggplot() +
  geom_boxplot(aes(y = fec, x = factor(year)),
               outliers = FALSE, show.legend = FALSE) +
  facet_wrap(~spat.unit, scales = "free_y") +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1, size = rel(.75))) +
  labs(title = "3SW")

```

# FIX

- 1,2 and 3 SW where relevant spawner ages exist. 
- combine 1,2 and 3 SW to one plot and add CI:s
